Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-24 main.tex
update
fixes
Update on Overleaf.
cool
quick fix
fix duplicate citations

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,journal,bibtex,acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen,review,anonymous]{acmart}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2024}
\acmYear{2024}
\acmDOI{XXXXXXX.XXXXXXX}

@@

\ifthenelse{1=0}{
\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
\newcommand\kyle[1]{\textcolor{olive}{Kyle: #1}}
\newcommand\teo[1]{\textcolor{red}{teo: #1}}
\newcommand\willow[1]{\textcolor{purple}{Willow: #1}}
\newcommand{\saman}[1]{\textcolor{orange}{Saman: #1}}
\newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
\newcommand\changwan[1]{\textcolor{blue}{CW: #1}}
}{
\newcommand{\TODO}[1]{}
\newcommand{\help}[1]{}
\newcommand{\saman}[1]{}
\newcommand{\kyle}[1]{}
\newcommand{\teo}[1]{}
\newcommand{\willow}[1]{}
\newcommand{\leftoff}{}
}

@@

\ccsdesc[500]{Software and its engineering~Control structures}
\ccsdesc[500]{Software and its engineering~Data types and structures}
\ccsdesc[300]{Software and its engineering~Imperative languages}
\ccsdesc[500]{Mathematics of computing~Mathematical software}
%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Array, Sparse, Structure, Control Flow, Language, Compiler}

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
\input{sections/intro}
\input{sections/background}
\input{sections/tensors}
\input{sections/lang}
\input{sections/compiler}
\input{sections/eval}
\input{sections/related}
\input{sections/conclusion}
%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
    To Mateusz, Hameer, and Jaeyeon for their excellent programming contributions to the Finch codebase.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{FinchOOPSLAWillow.bib, FinchOOPSLAOverleaf.bib}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Frequently Requested Finch kernels}

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,journal,bibtex,acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen,nonacm]{acmart}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{none}
\copyrightyear{2024}
\acmYear{2024}
\acmDOI{XXXXXXX.XXXXXXX}

@@

\ifthenelse{1=0}{
\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
\newcommand\kyle[1]{\textcolor{olive}{Kyle: #1}}
\newcommand\teo[1]{\textcolor{red}{teo: #1}}
\newcommand\willow[1]{\textcolor{purple}{Willow: #1}}
\newcommand{\saman}[1]{\textcolor{orange}{Saman: #1}}
\newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
\newcommand\changwan[1]{\textcolor{blue}{CW: #1}}
}{
\newcommand{\TODO}[1]{}
\newcommand{\help}[1]{}
\newcommand{\saman}[1]{}
\newcommand{\kyle}[1]{}
\newcommand{\teo}[1]{}
\newcommand{\willow}[1]{}
\newcommand{\changwan}[1]{}
\newcommand{\leftoff}{}
}

@@

\ccsdesc[500]{Software and its engineering~Control structures}
\ccsdesc[500]{Software and its engineering~Data types and structures}
\ccsdesc[300]{Software and its engineering~Imperative languages}
\ccsdesc[500]{Mathematics of computing~Mathematical software}
%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Sparse Array, Structured Array, Control Flow, Programming Language}

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
\input{sections/intro}
\input{sections/background}
\input{sections/tensors}
\input{sections/lang}
\input{sections/compiler}
\input{sections/eval}
\input{sections/related}
\input{sections/conclusion}
%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
  Intel and NSF PPoSS Grant CCF-2217064; DARPA PROWESS Award HR0011-23-C-0101; NSF SHF Grant CCF-2107244
\end{acks}

\clearpage

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{FinchOOPSLAWillow.bib}

%%
%% If your work has an appendix, this is the place to put it.
%\appendix
%\include{sections/appendix}

### 2024-04-24 sections/compiler.tex
fixes

\subsection{Dimensionalization}
    Looplets typically require the dimension of the loop extent to match the dimensions of the tensor. 
    %
    However, it is cumbersome to write the dimensions
    in loop programs, and most tensor compilers have a means of specifying the dimensions automatically.
    %
    In many pure Einsum languages like TACO, determining dimensions
    is not needed because any tensor dimensions that share an index are assumed to be the same~\cite{kjolstad_tensor_2017}.
    %
    Other languages, such as Halide, perform bounds inference where
    known bounds are symbolically propagated to fill in unknown bounds, often from output/input sizes to intermediates via some approximation such as interval analysis or polyhedral methods~\cite{ragan-kelley_halide_2013, grosser2012polly}.
    %
    We refer to the process of discovering suitable
    dimensions as \textbf{dimensionalization}. 
    %

>>>>

\subsection{Dimensionalization}
    Looplets typically require the dimension of the loop extent to match the dimensions of the tensor. 
    %
    However, it is cumbersome to write the dimensions
    in loop programs, and most tensor compilers have a means of specifying the dimensions automatically.
    %
    In many pure Einsum languages like TACO, determining dimensions
    is not needed because any tensor dimensions that share an index are assumed to be the same~\cite{kjolstad_tensor_2017}.
    %
    Other languages, such as Halide, perform bounds inference where
    known bounds are symbolically propagated to fill in unknown bounds, often from output/input sizes to intermediates via some approximation such as interval analysis or polyhedral methods~\cite{ragan-kelley_halide_2013,grosser_pollyperforming_2012}.
    %
    We refer to the process of discovering suitable
    dimensions as \textbf{dimensionalization}. 
    %

### 2024-04-24 sections/eval/graphs.tex
fixes

Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
%
Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier. 
%
The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway. 
%
Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.

Figure~\ref{fig:graph_result} compares performance to Graphs.jl, a Julia library, and the LAGraph Library, which implements graph algorithms with sparse linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
%
For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs. 
%
Although GraphBLAS uses hardwired optimizations, Finch is competitive. 
%
On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53). 
%
We did not include GAP-road as it timed out. %speedup is arith-mean
%
In Appendix B, we display the code for BFS and Bellman-Ford in Finch (57 and 50 LOC) and LAGraph (215 and 227 LOC), and invite readers to compare the clarity of the algorithms.

>>>>

Direction-optimization~\cite{beamer_direction-optimizing_2012} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
%
Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier. 
%
The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway. 
%
Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.

Figure~\ref{fig:graph_result} compares performance to Graphs.jl, a Julia library, and the LAGraph Library, which implements graph algorithms with sparse linear algebra using GraphBLAS~\cite{mattson_lagraph_2019}.
%
For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs. 
%
Although GraphBLAS uses hardwired optimizations, Finch is competitive. 
%
On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53). 
%
We did not include GAP-road as it timed out. %speedup is arith-mean
%
In Appendix B, we display the code for BFS and Bellman-Ford in Finch (57 and 50 LOC) and LAGraph (215 and 227 LOC), and invite readers to compare the clarity of the algorithms.

### 2024-04-24 sections/intro.tex
fixes

Our world is full of structured data.
%
Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
%
Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g., higher order FEMs) as well as in intermediate structures in many linear solvers (e.g., GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
%
In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
%
Even complex operators can be expressed as structured arrays.
%
For example, a convolution with a filter can be expressed as a matrix multiplication
with the Toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.

%
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
%
Special-purpose
compilers like TACO~\cite{kjolstad_tensor_2019}, Taichi~\cite{hu_taichi_2019}, StructTensor~\cite{ghorbani2023compiling}, or CoRa~\cite{fegade_cora_2022} which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions.
%
This trade-off is visualized in Tables \ref{tab:features} and \ref{tab:data_structures}.
%

@@

%
Prior implementations are incomplete because the abstractions they use are tightly coupled with the specific data structures that they support.
%
For example, TACO merge lattices represent Boolean logic over sets of non-zero values on an integer grid~\cite{kjolstad_tensor_2017}.
%
The polyhedral model allows various compilers to represent dense computations on affine regions~\cite{grosser2012polly}.
%
Taichi enriches single static assignment form with a specialized instruction for accessing only a single sparse structure, but it supports more control flow ~\cite{hu_taichi_2019}.
These systems tightly couple their control flow to narrow classes of data structures to avoid the challenges that occur when we intersect complex control flow with structured data. There are two challenges:
%writing efficient code over structured data

@@

\textbf{Performance on structured data is highly algorithm dependent}: The landscape of implementation decisions is dramatically unpredictable. 
%
For example, the asymptotic performance of sparse matrix multiplication can be impacted by the distribution of nonzeros, the sparse format, and the loop order~\cite{ahrens2022autoscheduling, zhang2021gamma}. 
%For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
This means that performance engineering for such kernels requires the exploration of a large design space, changing the algorithm as well as the data structures.

@@

\begin{enumerate}
\item More complex array structures than ever before. We are the first to extend level-by-level hierarchical descriptions to capture banded, triangular, run-length-encoded, or sparse datasets, and any combination thereof.
%
We have chosen a set of level formats that completely captures all combinations of relevant structural properties (zeros, repeated values, and/or blocks).
%
Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level. 
%
\item A rich structured array programming language with for-loops and complex control flow constructs at the same level of productivity of dense arrays. 
%
To our knowledge, the Finch programming language is the first to support if-conditions, early breaks, and multiple left hand sides over structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands.
%
\item A compiler that specializes programs to data structures automatically, facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures. Finch reliably utilizes four key properties of structure, such as structural zeros, repeated values, clustered non-zeros, and singletons.
%
\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs that we implement in this work.
%
For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{won2024continuous}.
%
\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as SpMV ($2.51\times$) and SpGEMM ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $4\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>

>>>>

Our world is full of structured data.

Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankar_petsc_nodate,bell_lessons_2007,mcauley_hidden_2013,balay_petsc_2020}.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)~\cite{shi_column_2020,golomb_run-length_1966}.
%
Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g., higher order FEMs) as well as in intermediate structures in many linear solvers (e.g., GMRES)~\cite{burns_dedalus_2020,saad_iterative_2003,oleary_scientific_2009}.
%
In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao_monarch_2022}.
%
Even complex operators can be expressed as structured arrays.
%
For example, a convolution with a filter can be expressed as a matrix multiplication
with the Toeplitz matrix of all the circular shifts of the filter~\cite{sze_efficient_2017}.

%
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{bradski_opencv_2000,anderson_lapack_1999,virtanen_scipy_2020,psarras_linear_2022}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{buluc_design_2017,mattson_lagraph_2019}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser_pollyperforming_2012}.  
%
Special-purpose
compilers like TACO~\cite{kjolstad_tensor_2019}, Taichi~\cite{hu_taichi_2019}, StructTensor~\cite{ghorbani_compiling_2023}, or CoRa~\cite{fegade_cora_2022} which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions.
%
This trade-off is visualized in Tables \ref{tab:features} and \ref{tab:data_structures}.
%

@@

%
Prior implementations are incomplete because the abstractions they use are tightly coupled with the specific data structures that they support.
%
For example, TACO merge lattices represent Boolean logic over sets of non-zero values on an integer grid~\cite{kjolstad_tensor_2017}.
%
The polyhedral model allows various compilers to represent dense computations on affine regions~\cite{grosser_pollyperforming_2012}.
%
Taichi enriches single static assignment form with a specialized instruction for accessing only a single sparse structure, but it supports more control flow ~\cite{hu_taichi_2019}.
These systems tightly couple their control flow to narrow classes of data structures to avoid the challenges that occur when we intersect complex control flow with structured data. There are two challenges:
%writing efficient code over structured data

@@

\textbf{Performance on structured data is highly algorithm dependent}: The landscape of implementation decisions is dramatically unpredictable. 
%
For example, the asymptotic performance of sparse matrix multiplication can be impacted by the distribution of nonzeros, the sparse format, and the loop order~\cite{ahrens_autoscheduling_2022,zhang_gamma_2021}. 
%For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
This means that performance engineering for such kernels requires the exploration of a large design space, changing the algorithm as well as the data structures.

@@

\begin{enumerate}
\item More complex array structures than ever before. We are the first to extend level-by-level hierarchical descriptions to capture banded, triangular, run-length-encoded, or sparse datasets, and any combination thereof.
%
We have chosen a set of level formats that completely captures all combinations of relevant structural properties (zeros, repeated values, and/or blocks).
%
Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou_format_2018,  hu_taichi_2019, strout_sparse_2018, bernstein_ebb_2016} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level. 
%
\item A rich structured array programming language with for-loops and complex control flow constructs at the same level of productivity of dense arrays. 
%
To our knowledge, the Finch programming language is the first to support if-conditions, early breaks, and multiple left hand sides over structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands.
%
\item A compiler that specializes programs to data structures automatically, facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures. Finch reliably utilizes four key properties of structure, such as structural zeros, repeated values, clustered non-zeros, and singletons.
%
\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs that we implement in this work.
%
For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{anonymous_authors_continuous_2024}. Finch is also used as a compiler backend for the Python PyData/Sparse library \cite{abbasi_plans_2023}.
%
\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as SpMV ($2.51\times$) and SpGEMM ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $4\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
